# Analysis of longitudinal data

```{r} 
# set figure sizes for knitting
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
```
Libraries to load up.
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```



First we load up the datasets. We will also need the original RATS data in one of the analyses.
```{r}
BPRSL = read.csv("BPRSL.csv", as.is = TRUE, header = TRUE, row.names = NULL)

RATSL = read.csv("RATSL.csv", as.is = TRUE, header = TRUE, row.names = NULL)

RATS = read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = T, sep = '\t')
```

Let's check the datasets
```{r}
str(BPRSL)
str(RATSL)
```
Seems we need to refactor the factorial variables again, we'll recheck with **str** to see that it worked again.
```{r}
# use as.factor to transform them back to factorial variables in both datasets and confirm with str
BPRSL$treatment = as.factor(BPRSL$treatment)
BPRSL$subject = as.factor(BPRSL$subject)

RATSL$ID = as.factor(RATSL$ID)
RATSL$Group = as.factor(RATSL$Group)
str(BPRSL)
str(RATSL)
```


Now that we have our data set, let's start with the analysis. We'll begin with the RATSL data first by summarizing it.
```{r}
summary(RATSL)
```


We have our five variables needed for the analysis. We'll start with an graphical plot.
```{r}
# we use ggplot to creat the parallel plots so that each group of rat is separated for better visualization
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

This plot produces the weight curves during the time for each group separately. From the plot it would be suggested that rats in group 1 gain the least weight. We can also try if it changes with standardization of the dataset.
```{r}
RATSL = RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
```
```{r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```

We can also plot the mean errors to these. First we create a new summarized data with the Weight and n = 11 times of measurement spots in the RATS data (week1-week64, measurement once every week).
```{r}
n = 11
  
RATSS = RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
```

Let's take glimpse of the data.
```{r}
glimpse(RATSS)
```


We now have the mean values and the standard errors with 11(remember the amount of measures for each rat) x 3 groups = 33 observations (11 for each of the group). 


We are now ready to plot this summary data.
```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=2.5) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9, 0.4)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```


Groups 2 and 3 had the least amount of different rats so their standard errors are large, group 2 also had an outlier rat with greater values than other rats in it so the standard errors end up even larger than in the group 3. In general though group 2 and 3 rats seem to end up gaining weight more than group 1 rats although the high standard errors for group make it a bit harder to interpt straight from the plot.

We can do a new summarising without the first week starting measurement.
```{r}
RATSL10S = RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
```

Then let's glimpse the data.
```{r}
glimpse(RATSL10S)
```

So now we have the mean weights for weeks 2-11 and the ID's as well as th rat groups. We can now boxplot it for outliers.
```{r}
ggplot(RATSL10S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 2-11")
```


The outliers here seem to be below 250 and above 550. For group 3 there is also an outlier at 500 weight though it is in the midde of the dataset even if its seen as an outlier for group 3. We can however use the 250 and 550 as weight limits and remove the ID = 13 outlier from group 3.
```{r}
RATSL10S1 = RATSL10S %>%
  filter(mean > 250 & mean < 550 & ID != 13)
```

We have now removed the 3 outlier variables from the data. Let's perform a chisquare-test for dataset now comparing the groups (t-test is now not possible as the grouping variable has 3 groups).


We can do qqplot for normality test for the mean-variable in the RATSL10S1 with the ggbur-package using a density function. For normally distibuted we should see a bell-shaped curve.

```{r}
library("ggpubr")
par(mfrow = c(1,1))
ggdensity(RATSL10S1$mean, fill = "lightgray")
```


This definitely does not seem to be normally distributed. However wee can do a non-parametric test with Kruskal-Wallis on the data
```{r}
kruskal.test(mean ~ Group, data = RATSL10S1)
```

This yet again indicates a difference between the group means.

We can do further analysis by adding back the baseline and looking whether there is difference in the growth of weight with a linear model.
```{r}
RATSL10S2  = RATSL10S %>%
  mutate(baseline = RATS$WD1)
```


```{r}
#linear model for weight difference
fit.rat = lm(mean ~ baseline + Group, data = RATSL10S2)
summary(fit.rat)
```

It would seem that the groups do not quite differ in the weight gain after all even though they differ between each other in the baseline as could already be seen the in the plots.

Next we can move onto the blood pressure data(BPRS).
```{r}
BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRSL_reg)
```

It would seem that in this model the bprs differs between week but not between the treatment groups. The explanation of the variance of blood pressure is only 18% as well.

We can make an intercept model with the lmer tool in lme4-package.
```{r}
library("lme4")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```
Let's make a random slope model.
```{r}
BPRS_ref1 = lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
```
We can check the difference with the ANOVA test.
```{r}
anova(BPRS_ref1,BPRS_ref)
```

There does seem to be a difference between the random intercept and random slope models.

Let's do yet another model, interaction model this time.
```{r}
BPRS_ref2 = lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
```
Now we can do an ANOVA test on these two models to see if it gets better.
```{r}
anova(BPRS_ref2, BPRS_ref1)
```

There doesn't seem to be a difference between these two models, which suggest that the interaction term doesn't really improve the situation here, so we can proceed with the ref1 model.

We will create a fitted blood pressure vector with fitted tool used on the ref1 model.
```{r}
BP.fit = fitted(BPRS_ref1)
```

Let's put the fitted data back into BPRSL data.
```{r}
BPRSL = BPRSL %>%
  mutate(BP.fit)
```

Plot the original blood pressure variable.
```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```


It is difficult to say from this whether they differ.


Now to compare, plot of the fitted blood pressure variable.
```{r}
ggplot(BPRSL, aes(x = week, y = BP.fit, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$BP.fit), max(BPRSL$BP.fit)))
```

The fitted models plots definitely don't seem to differ from each other. We can say with some confidence that the blood pressures did not differ between the treatments or treatments are equally good. The result is also inline with the swapped results you can get from the datacamp.

```{r}
date()
```

