# Regression and model validation

This week we have the learning 2014 data, already changed with the script found in https://github.com/JuusoPJ/IODS-project

```{r}
# read file made with script
students2014 = read.csv("learning2014.csv", as.is = TRUE, header = TRUE)

# remove the first x index column
students2014 = students2014[c(2,3,4,5,6,7,8)]
```

```{r}
str(students2014)
```
The data consists of 166 observations with 7 variables.
```{r}
# access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)
```
```{r}
# create a more advanced plot matrix with ggpairs()
p = ggpairs(students2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```


```{r}
# load corrplot package to plot correlations between all the variables (similar but far less heavy than ggplot)
library(corrplot)
```
```{r}
# create vector for just numeric variables
x = students2014[, c("age", "attitude", "deep", "stra","surf","points")] 
# correlation matrix
cor(x, use = "pairwise") 
```
```{r}
# graphical representation of correlations from corrplot package
corrplot.mixed(cor(x, use = "pairwise"), order ="hclust") 
```
```{r}
# linear model with three most correlated variable
lm.p = lm(points ~ attitude + surf + stra, data = students2014)
summary(lm.p)
```
The model explains 20% of the variance of points, with only attitude having a significant p-value and so differing from null hypothesis.
```{r}
# new model with the only significant effect
lm.pn = lm(points ~attitude, data = students2014) 
summary(lm.pn)
```
The model now explains 19% of the variance of points with points increasing 0.35 for every 1 increase in attitude.
```{r}
# scatter plot with the linear model line
qplot(attitude, points, data = students2014) + geom_smooth(method = "lm") 
```
```{r}
# plot to 2x2 area
par(mfrow = c(2,2)) 

# automatically generates 4 typically used diagnostic plots
plot(lm.pn) 
```

1. Residuals do not show a pattern where the distribution of residuals varies depending on fitted values.
2. QQ-plot seems to be nicely on the line making the SEs and P-values of the model coefficient reliable.
3. Standardized residuals vs fitted should show a similar magnitude along y-axis of each region on the x-axis, which it does.

```{r}
date()
```
